\chapter{Maintenance}\label{ch:maintain}

\section{New versions, patches}

New versions will be announced on the Dalton homepage
(\verb|http://daltonprogram.org|)
and the \dalton\ mailing-list
(\verb|dalton-users@kjemi.uio.no|)\index{mailing list}.

In between releases, bug fixes will be distributed as
patches\index{patches} to an
existing version. New patches will be
freely available from the 
Dalton homepage, and will be announced on the dalton-users mailing
list. Patches will normally be distributed in order to correct
significant errors. In case of minor changes, explicit
instructions on modifying the source file(s) may be given.

In general, users should not make local
modifications in the source code, as this usually makes it
much harder to incorporate the changes into subsequent versions.
It is more convenient to use the C preprocessor code.  Indeed, by
judicious use of local \verb|#define|\index{define} variables (this
is described in any book on C that also describes the C
preprocessor) it is often possible to isolate local
modifications\index{modifications} completely, making it much easier
to apply them to later
releases.

\section{Reporting bugs and user support}

The \lsdalton\ program is distributed to the computational chemistry
community with no obligations on the side of the
authors. The authors thus take no responsibility\index{responsibility}
for the performance
of the code or for the correctness\index{correctness} of the
results. This distribution
policy gives the authors no responsibility for problems experienced by
the users when using the \lsdalton\ program.

A mailing-list\index{mailing list} ---
\verb|dalton-users@kjemi.uio.no| --- has been created for
the purpose of communication among the authors and the users about
new patches and releases, and communication between users of the
program. The authors hope that the
users will be able to help each others out when problems
arise. All authors of the \lsdalton\ program also receive mail from the
mailing-list, but we do not guarantee that any author will reply to
requests posted to the mailing-list. By default, all users signing a
license agreement form will be added to this mailing list. If more
people use the same licensed version, these users may send mail to
\verb|dalton-admin@kjemi.uio.no| asking to be added to this list as
well as indicating whose licensed version of the program they are using.

Bug reports\index{bugs} are to be reported to \verb|dalton-admin@kjemi.uio.no| and
will be dealt with by one of the authors, although no responsibility
on the promptness of this response is given. In general, serious bugs
that have been
reported and fixed will lead to a new patch\index{patches} of the program, distributed from the
Dalton homepage (\verb|http://daltonprogram.org|).
