
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Excited state linear response functions and
         two-photon transition moments between two excited states:
         \Sec{CCEXLR}}
\label{sec:ccexlr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{Coupled Cluster!linear response}
\index{linear response!excited states}
\index{response!linear}
\index{excited state!linear response}
\index{polarizability!frequency-dependent}
\index{two-photon!transition moment!excited states}
\index{second residues of cubic response functions}
\index{second-order properties!excited states}
\index{properties!second-order}

In the \Sec{CCEXLR} section input that is specific for 
double residues of coupled cluster cubic response functions 
is read in.
Results obtained using this functionality should cite
Ref.~\cite{Haettig:EXCITED,Haettig:EXLR}. 
This section includes:
\begin{itemize}
\item frequency-dependent second-order properties of excited states
      $$ \alpha^{(i)}_{AB}(\omega) = 
         -\langle\langle A; B\rangle\rangle^{(i)}_\omega $$
      where $A$ and $B$ can be any of the one-electron operators
      for which integrals  are available in the \Sec{*INTEGRALS}
      input part.
\item two-photon transition moments between excited states.
\end{itemize}
Double residues of coupled cluster cubic response functions are
implemented for the models CCS, CC2, and CCSD.
%Publications that report results obtained with double residues
%of CC cubic response functions should cite Ref.\ \cite{Haettig:EXLR}.

\begin{center}
\fbox{
\parbox[h][\height][l]{12cm}{
\small
\noindent
{\bf Reference literature:}
\begin{list}{}{}
\item C.~H\"{a}ttig, and P.~J{\o}rgensen \newblock {\em J.~Chem.~Phys.}, {\bf 109},\hspace{0.25em}4745, (1998).
\item C.~H\"{a}ttig, O.~Christiansen, S.~Coriani, and P.~J{\o}rgensen \newblock {\em J.~Chem.~Phys.}, {\bf 109},\hspace{0.25em}9237, (1998).
\end{list}
}}
\end{center}

\begin{description}
\item[\Key{ALLSTA}] 
Calculate excited-state polarizabilities for all excited states.
 
\item[\Key{DIPOLE}] 
Evaluate all symmetry-allowed elements of the dipole--dipole tensor
of the double residues of the cubic response function 
(a maximum of six components for second-order properties, and a
 maximum of nine for two-photon transition moments).
 
\item[\Key{FREQ}]  \verb| |\newline
%  or \Key{FREQUE}
\verb|READ (LUCMD,*) MFREQ|\newline
\verb|READ (LUCMD,*) (BEXLRFR(IDX),IDX=NEXLRFR+1,NEXLRFR+MFREQ)|

Frequency input for $\alpha^{(i)}_{AB}(\omega)$.
 
\item[\Key{HALFFR}] 
Use half the excitation energy as frequency argument for two-photon
transition moments.
Note that the \Key{HALFFR} keyword is incompatible with a 
user-specified list of frequencies. \\
For excited-state second-order properties the \Key{HALFFR} keyword is
equivalent to the \Key{STATIC} keyword.
 
\item[\Key{OPERAT}] \verb| |\newline
\verb|READ (LUCMD,'(2A)') LABELA, LABELB|\newline
\verb|DO WHILE (LABELA(1:1).NE.'.' .AND. LABELA(1:1).NE.'*')|\newline
\verb|  READ (LUCMD,'(2A)') LABELA, LABELB|\newline
\verb|END DO|

Read pairs of operator labels.
For each of these operator pairs, the double residues of the cubic response
function will be evaluated at all frequencies.
Operator pairs which do not correspond to symmetry-allowed
combinations will be ignored during the calculation.
 
\item[\Key{PRINT}] \verb| |\newline
\verb|READ (LUCMD,*) IPRINT|

Set print parameter for the \Sec{CCEXLR} section.
 
\item[\Key{SELSTA}] \verb| |\newline
\verb|READ (LUCMD,'(A80)') LABHELP|\newline
\verb|DO WHILE(LABHELP(1:1).NE.'.' .AND. LABHELP(1:1).NE.'*')|\newline
\verb|  READ(LUCMD,*) ISYMS(1), IDXS(1), ISYMS(2), IDXS(2)|\newline
\verb|END DO|

Read symmetry and index of the initial state and the final state.
If initial and final state coincide one obtains excited state
second-order properties (or, more precisely, the difference of the excited state second-order property relative to the ground-state property), if the two excited states are different one obtains the
two-photon transition moments between the two excited states.
 
\item[\Key{STATIC}] 
Add $\omega = 0$ to the frequency list.
 
\item[\Key{USELEF}] 
Use left excited-state response vectors instead of the right excited-state response vectors (default is to use the right excited-state
response vectors).
 
\end{description}
